---
title: "Class18"
author: "Eric Wang A17678188"
format: pdf
toc: TRUE
---


## Background
Pertussis (a.k.a. whooping cough) is a common lung infection caused by the B. Pertussis. It is a bacterial disease. 

This can infect all ages but is most sever for those under 1 year of age. 

The CDC track the number of [reported cases in US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html):

We can "scrape" this data with the **datapasta** package

## Investigating pertussis cases by year

> Q1. Use ggplot to make a plot of cases numbers over time. 

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024L,2025L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538,21996)
       )
```

```{r}
head(cdc)
```

```{r}
library(ggplot2)

ggplot(cdc) + aes(year, cases) + geom_point() + geom_line() + labs(x="Year", y = "Number of cases", title = "Pertussis Cases by Year (1922-2025")
```

> Q. Add some jaor milestones including the first wP vaccine roll-out (1946), the switch to the newer aP vaccine 1996, the COVID years (2020)

```{r}
ggplot(cdc) + aes(year, cases) + geom_point() + geom_line() + labs(x="Year", y = "Number of cases", title = "Pertussis Cases by Year (1922-2025") + geom_vline(xintercept=1946, col="lightblue", lty=2)+ 
geom_vline(xintercept=1996, col="lightyellow", lty=2) +
  
  geom_vline(xintercept=2020, col= "green",lty = 2)
```

There were high case numbers in the pre 1940s, they were reduced dramatically due to the introduction of the first vaccine. After the the aP was introduced but it worked differently from wP. It introduced protection but it showed that the protect wore off faster than wP. 

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

## CMI-PB project
[Computational Models of Immunity - Pertussis Boost](https://www.cmi-pb.org/) project aims to provide the scientific community with this very information.

They make their data available via JSON format returning API. We can read this in R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
                     
head(subject)
```

> Q. How many wP and aP individuals are in this `subject` table? 

```{r}
table(subject$infancy_vac)
```

> Q. What is the biolgoical sex breakdown?

```{r}
table(subject$biological_sex)
```

> Q. What is the breakdown of race and biological sex

```{r}
table(subject$race, subject$biological_sex)
```

Let's read some more databse tables: 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)

ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To analyze this data we need to first "join"(merge/link) the different tables so we have all the data in one place not spread across different tables. 

We can use the `*_join()` family of functions from **dplyr** to do this. 

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
head(meta)
```

> Q. What antibody isotypes are measured for these patients?

```{r}
table(abdata$isotype)
```

>Q> What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```
It increased compared to the previous two years. 


>Q. What antigens are reported?

```{r}
table(abdata$antigen)
```


Let's focus on the IgG isotype and make a plot of MFI_normalized for all antigens. 

```{r}
igg <- abdata %>% filter(isotype=="IgG")
head(igg)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

> Q. Is there a differrence for aP vs wP individuals with these values

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot()
```


> Q. Is there a temporal response - i.e. do values increase or decrease overtime? 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() + facet_wrap(vars(visit))
```

## Focus on "PT" Pertusisis Toxin antigen

```{r}
pt.igg.21 <- igg |> filter(antigen == "PT",
              dataset== "2021_dataset")
```


```{r}
ggplot(pt.igg.21) + aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) + geom_point() + geom_line() + geom_vline(xintercept = 14, col = "darkgreen", lty=2)
```

```{r, warning=FALSE}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")
abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() + geom_smooth(
  data = abdata.21 %>% 
           filter(isotype == "IgG", antigen == "PT"),
  mapping = aes(
      x = planned_day_relative_to_boost,
      y = MFI_normalised,
      color = infancy_vac
  ), se = FALSE,
  linewidth = 1.5,
  inherit.aes = FALSE, span = 0.4
)+
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```




